---
title: "Basics: Building a module"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`outsider` allows users to install and run programs on their own computer
without the need to leave the R environment. If there is a command-line program
that is not available through `outsider` you can create your own module! If you
are able to install the program on your own machine and have some experience
with R packages and GitHub then you should be readily able to create one.
In this article, we will outline what an `outsider` module is and then provide
a simple walkthrough for creating the simplest of simple modules: a command line
program that prints whatever text the user provides it.

To follow this guide you will need the following:
*A GitHub account (sign-up [here](https://github.com/))
*A Docker-Hub account (sign-up [here](https://hub.docker.com/))

# Basics

## Module structure

At its heart, an `outsider` module is just an R package with a `Dockerfile`.
All modules have the following basic file/folder structure:

```
# Core files and folders of a module #
- DESCRIPTION
- R/
    - functions.R
- dockerfiles/
    - latest/
        - Dockerfile
- om.yaml
- README.md
- .travis.yml
```

*R files and folders*

The R package is encapsulated by the `DESCRIPTION` file and the `R/` folder.
The first of these describes the package details (package name, author,
dependencies, etc.) and the second contains the R code that make up the package.
The `R/` folder can have any number of scripts with whatever names a developer
chooses (skeleton default `functions.R`).

*Docker files and folders*

The `outsider` package and its modules depend on Docker to run. Docker acts
like a virtual machine on a user's computer. It allows users to run specific
code, often designed for different operating systems, on any machine that has
Docker installed. These code snippets are called Docker images. And these
images are described by a `Dockerfile`. The `dockerfiles/` folder in the
module contains the `Dockerfile` that describes how to install the external
command-line program that the develop wishes to run through R. `dockerfiles/`
can have multiple versions of a `Dockerfile` but every module must have a
`latest/`.

*GitHub files and folders*

To make modules discoverable on GitHub, all modules require an `om.yml`.
This file has two elements (program and details) encoded in the
[YAML format](https://en.wikipedia.org/wiki/YAML). In addition, the module
has a `README.md` file that provides the text describing the module on the
module's GitHub homepage and a `.travis.yml` that provides instructions
detailing how the module should be tested on
[Travis-CI](https://en.wikipedia.org/wiki/Travis_CI).

# Walkthrough

We will walk you through how to create your very own `outsider` module that will
simply print (through `echo`) any text provided. This process comes in a
series of steps: 1) generate core files and folders, 2) create Docker image, 3)
document and build R package, 4) try the module, 5) upload to GitHub and Docker
Hub.

## Step 1: Generate the files and folders

As displayed above, we need to generate the core files and folders of a module.
This process can be easily performed using the `.module_skeleton()` function.
This function takes a few details about the developer and the program and then
generates all the necessary core file structures.

> **What's with the `.` at the beginning of functions?** Dots at the
beginning of an object in R are hidden. To prevent the average user from running
`outsider` functions intended for use by module developers a `.` is added.

The necessary information required for the function to run are our GitHub and
Docker Hub usernames plus the name for the program we wish to provide as a
module (which is `echo` -- a UNIX command for printing). In these code snippets,
the usernames are of the `outsider` maintainer. In order for these examples to
work for you, you will need to change "dombennett" to your own usernames.
(Note, your usernames may differ for GitHub and Docker Hub).

```{r skeleton}
library(outsider)
.module_skeleton(github_user = 'dombennett', program_name = 'echo',
                 docker_user = 'dombennett', flpth = getwd())
```

The above code will create an `outsider` module with the name `om..echo`.
After running the above code, you should take a minute to inspect the generated
files. In particular you should look at the `DESCRIPTION` file, the `Dockerfile`
and `functions.R`.

> **Why `om..echo`?** All modules must start with "om.." in order for them to
be discovered on GitHub.

At this stage, we would then edit the `dockerfiles/latest/Dockerfile` and the
`R/functions.R` to work for our chosen external program. But because `echo`, the
program we wish to port through `outsider`, is so simple we don't actually have
to make any changes to these files. By default, the `Dockerfile` is based on
Ubuntu which ships with `echo` and our starting function in `functions.R` is
based around running `echo`: it parses the arguments, creates an outsider
object, and then launches the object.

```{r echo-function, echo=TRUE, eval=FALSE}
# The echo function in om..echo
echo <- function(...) {
  # convert the ... into an argument list
  arglist <- outsider::.arglist_get(...)
  # create an outsider object: describe the arguments and program
  otsdr <- outsider::.outsider_init(repo = 'dombennett/om..echo',
                                    cmd = 'echo',
                                    arglist = arglist)
  # run the command
  outsider::.run(otsdr)
}
```

> **What's `...`?** In function calls in R, `...` indicate that any number of
arguments can be provided to a function. The `.arglist_get` function will take
the `...` and convert them into a character vector that can be parsed.
`outsider` recommends module functions make use of this feature so that any
number of arguments can be passed to external programs. Additionally, the
developer would then not need to document all the arguments of the external
program. For many external programs there may be hundreds of arguments, all of 
which are likely to be already documented; viewable through commands like
`-h` or `--help`.

## Step 2: Create the Docker image

> **Dockerfile commands** Dockerfiles are series of instructions for
constructing a virtual machine called a Docker image. Each command begins with
a capitalised instruction followed by arguments. The most common instruction
would be "RUN", this executes command-line code within the Docker image system.
For example, `RUN echo "command!"` would pass "command!" to the program "echo".
All Dockerfiles begin with a FROM instruction, which pulls a Docker image on
which to build your own image. For example, many images are built upon the Linux
operating system Ubuntu in which case the first line of the Dockerfile would
be "FROM ubuntu:latest". This first line would then download the latest Ubuntu
Docker image and all subsequent "RUN" instructions would be running in Ubuntu
command-line. For far more detailed information on Dockerfiles, see the Docker
docs.

In our `om..echo` we have a dockerfiles folder that contains a Dockerfile
describing the Docker image for our `echo` program. Our "latest" Dockerfile
contains the instructions to pull the Docker image of the latest Ubuntu release,
to create a folder called "working_dir" and then set this new folder as the
"WORKDIR".

```
# Example host distro
FROM ubuntu:latest

# Install program using RUN lines

# outsider *requires* working_dir
RUN mkdir /working_dir
WORKDIR /working_dir
```

> **What's the WORKDIR?** The WORKDIR sets the starting directory when a command
is passed to the Docker image. All `outsider` modules require this to be
"working_dir" as it allows `outsider` functions to know where to transfer files
to and from the virtual machine.

Using our Dockerfile, we now need to build our Docker image. This requires
running commands via command-line. Before continuing, open your favourite
terminal or command-prompt program and type `docker --help` to ensure that your
Docker program is up and running. The command to then build a docker image from
a Dockerfile is the following:
`docker build -t [docker-hub-name]/[docker_image]:[tag] [Dockerfile directory]`
Where the `-t` indicates the tagged name of the Docker image and the final
argument is the file path to the location of the Dockerfile. Note that for a
module, the GitHub repo name has `..` whereas the Docker image name has `_`.
In our example, we can run the following command in a terminal:

```{bash docker-build}
# docker build -t [tagged name] [directory containing Dockerfile]
# remember to change 'dombennett' to your Docker Hub
docker build -t dombennett/om_echo:latest om..echo/dockerfiles/latest/.
```

> **What is a tag?** A Docker tag is akin to the version number of the image.
By default, if no tag is provided,  Docker will use 'latest'.

## Step 3: Document and build R package

If the Docker image is successfuly built, we can now install the R package. To
do this we can run somme functions from within R using the [`devtools`](https://github.com/r-lib/devtools) package.

```{r build-package}
# first document
devtools::document('om..echo')
# then install
devtools::install('om..echo')
```

> **What does `document()` do?** In addition to the core files indicated above,
an R package also requires R documentation files that are stored in `man/` --
these provide the `?[function]` utility. The `document()` function generates
these files via [`roxygen`](https://github.com/klutometis/roxygen) comments and
tags located in the R scripts that make the package, i.e. all comments that
begin `#'`. For more information, look up "object documentation in R".

## Step 4: Try the module.

With a Docker image and an installed R package we are now ready to try out the
module before we upload online.

```{r test-package}
library(outsider)
# the repo always refers to the future github repo
echo <- module_import('echo', repo = 'dombennett/om..echo')
echo('hello world!')
```

## Step 5: Upload to GitHub and Docker Hub

After we have played with the module and ensured it works as we would hope we
can upload it to our GitHub and Docker accounts so that others may download it.

**GitHub**

First create a repository on your online GitHub account by clicking '+', ensure
the remote repository has the same name as your module, i.e. `om..echo`. Then
initiate a local git repository and push it to your remote.

```{bash pushgit, eval=FALSE, include=TRUE}
git init
git remote add origin https://github.com/[GitHub-username]/om..echo
git push origin master
```

**Docker Hub**

For Docker the command is simpler: within the `om..echo` directory, run
`docker push` followed by `[Docker Hub username]/[Docker image name]:[tag]`

```{bash pustdocker, eval=FALSE}
docker push dombennett/om_echo:latest
```

## Next steps and clean-up

**Testing**

After upload we can then test the functionality of the package from an end-users
perspective by running `.module_test`. This function tries installing the
package using the online repos, it then tries to import all the module functions
and running their examples.

**Tips**

The best way to learn how to build your own module is to look at how others have
created modules for programs you are familiar with. For more detailed written
information on building modules, see the [Advanced: Building a module](https://antonellilab.github.io/outsider/articles/advanced_build.html)

**Delete it all**

Don't want `om..echo` on your computer? Delete it so ...

```{r clean-up}
# to delete the Docker image and uninstall the R package
module_uninstall(repo = 'dombennett/om..echo')
# to delete the repo folder
unlink(x = 'om..echo', recursive = TRUE, force = TRUE)
```
