---
title: "Run commands on a server"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = '')
```

`outsider` allows users to launch commands on a remote machine. This can be
handy if the user wishes to run their external programs on more powerful
machines. All that is required is that the remote server has
[Docker](https://www.docker.com/) and can be connected to via
[ssh](https://en.wikipedia.org/wiki/Secure_Shell). To then run `outsider`
module programs on the remote machine, an `ssh` connection must be set up
using the `ssh_setup` function and the [`ssh`](https://github.com/ropensci/ssh)
package. Once set-up, all external programs will be run on the external
machine, not the local, with all input and output files being automatically sent
and retrieved by `outsider`.

```{r eval=FALSE}
# Libs
library(ssh)
library(outsider)

# SSH
session <- ssh_connect(host = "username@ip.address:port")
ssh_setup(session)

# Run
repo <- 'dombennett/om.hello.world'
if (!is_module_installed(repo = repo)) {
  module_install(repo = repo)
}
module_import(repo = repo, fname = 'hello_world')
# hello world will be run on the remote machine
hello_world()

# Clean up
ssh_teardown()
```

